---
title: "Vancouver Cherry Blossoms — Street Analysis"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(sf)
library(dplyr)
library(mapgl)
library(DT)
library(here)
```

```{r}
#| label: load-data
#| message: false

streets_sf     <- st_read(here("data/processed/streets_counted.geojson"), quiet = TRUE)
street_summary <- read.csv(here("data/processed/street_summary.csv"))
```

## Streets by total cherry blossom count

Streets with more trees are darker red. Grey lines are streets with no cherry blossoms.

```{r}
#| label: map-count

all_streets <- st_read(here("data/final/streets.geojson"), quiet = TRUE)

maplibre(style = openfreemap_style("bright"), bounds = streets_sf) %>%
  add_line_layer(
    id           = "all-streets",
    source       = all_streets,
    line_color   = "#dddddd",
    line_width   = 1,
    line_opacity = 0.4
  ) %>%
  add_line_layer(
    id         = "streets-count",
    source     = streets_sf,
    line_color = interpolate(
      column = "total_trees",
      values = c(1, median(streets_sf$total_trees), max(streets_sf$total_trees)),
      stops  = c("#fde0c5", "#f4895f", "#7f0000")
    ),
    line_width   = interpolate(
      column = "total_trees",
      values = c(1, max(streets_sf$total_trees)),
      stops  = c(1.5, 6)
    ),
    line_opacity = 0.9,
    tooltip      = "street_label",
    popup        = concat(
      "<b>", get_column("street_label"), "</b><br/>",
      "Trees: ", get_column("total_trees"), "<br/>",
      "Length: ", number_format(get_column("total_length"), 0), "m"
    )
  )
```

```{r}
#| label: table-count

street_summary %>%
  arrange(desc(total_trees)) %>%
  transmute(
    Street          = street_label,
    Type            = street_type,
    `Total trees`   = total_trees,
    `Length (m)`    = round(total_length),
    `Trees / 100m`  = round(density, 1),
    Segments        = n_segments
  ) %>%
  head(30) %>%
  datatable(
    caption  = "Top 30 streets by total cherry blossoms",
    rownames = FALSE,
    options  = list(pageLength = 30, dom = "t")
  )
```

## Streets by cherry blossom density (trees per 100m)

Density rewards short, packed blocks over long streets that happen to have many trees spread thin.

```{r}
#| label: map-density

# filter to streets with at least 5 trees to suppress stub-segment noise
streets_dense <- streets_sf %>% filter(total_trees >= 5)

maplibre(style = openfreemap_style("bright"), bounds = streets_dense) %>%
  add_line_layer(
    id           = "all-streets",
    source       = all_streets,
    line_color   = "#dddddd",
    line_width   = 1,
    line_opacity = 0.4
  ) %>%
  add_line_layer(
    id         = "streets-density",
    source     = streets_dense,
    line_color = interpolate(
      column = "density",
      values = c(0, median(streets_dense$density), max(streets_dense$density)),
      stops  = c("#edf8e9", "#74c476", "#00441b")
    ),
    line_width   = interpolate(
      column = "density",
      values = c(0, max(streets_dense$density)),
      stops  = c(1.5, 6)
    ),
    line_opacity = 0.9,
    tooltip      = "street_label",
    popup        = concat(
      "<b>", get_column("street_label"), "</b><br/>",
      "Trees / 100m: ", number_format(get_column("density"), 1), "<br/>",
      "Total trees: ", get_column("total_trees"), "<br/>",
      "Length: ", number_format(get_column("total_length"), 0), "m"
    )
  )
```

```{r}
#| label: table-density

top30_density <- street_summary %>%
  filter(total_trees >= 5) %>%
  arrange(desc(density)) %>%
  head(30) %>%
  mutate(rank = row_number())

top30_density %>%
  transmute(
    `#`             = rank,
    Street          = street_label,
    Type            = street_type,
    `Trees / 100m`  = round(density, 1),
    `Total trees`   = total_trees,
    `Length (m)`    = round(total_length),
    Segments        = n_segments
  ) %>%
  datatable(
    caption  = "Top 30 streets by cherry blossom density (min. 5 trees)",
    rownames = FALSE,
    options  = list(pageLength = 30, dom = "t")
  )
```

## Top 30 density streets — labelled map

Numbers correspond to the density ranking in the table above. Shaded blobs group streets that are within ~400m of each other.

```{r}
#| label: map-density-labelled

neighbourhoods <- st_read(here("data/original/local-area-boundary.geojson"), quiet = TRUE)

# Centroids of top 30 streets for label placement
top30_sf <- streets_sf %>%
  filter(total_trees >= 5) %>%
  arrange(desc(density)) %>%
  head(30) %>%
  mutate(
    rank       = row_number(),
    rank_label = as.character(rank)
  ) %>%
  st_centroid()

# Cluster nearby streets into blobs by buffering centroids and unioning overlaps.
# Project to UTM 10N (metres) for accurate buffering, then back to WGS84.
top30_clusters <- top30_sf %>%
  st_transform(32610) %>%
  st_buffer(500) %>% # change this to make the circles bigger
  st_union() %>%
  st_cast("POLYGON") %>%
  st_as_sf() %>%
  mutate(cluster_id = row_number()) %>%
  st_transform(4326)

maplibre(style = openfreemap_style("bright"), bounds = streets_dense) %>%
  # --- 1. Neighbourhood boundaries (bottom) ---
  add_fill_layer(
    id           = "neighbourhoods-fill",
    source       = neighbourhoods,
    fill_color   = "transparent",
    fill_opacity = 0,
    tooltip      = "name"
  ) %>%
  add_line_layer(
    id           = "neighbourhoods-borders",
    source       = neighbourhoods,
    line_color   = "#aaaaaa",
    line_width   = 1,
    line_opacity = 0.6,
    line_dasharray = c(2, 1)
  ) %>%
  # --- 2. Cluster blobs ---
  add_fill_layer(
    id           = "clusters-fill",
    source       = top30_clusters,
    fill_color   = "#fde68a",
    fill_opacity = 0.25
  ) %>%
  add_line_layer(
    id           = "clusters-border",
    source       = top30_clusters,
    line_color   = "#d97706",
    line_width   = 1.5,
    line_opacity = 0.7
  ) %>%
  # --- 3. All streets (context) ---
  add_line_layer(
    id           = "all-streets",
    source       = all_streets,
    line_color   = "#dddddd",
    line_width   = 1,
    line_opacity = 0.4
  ) %>%
  # --- 4. Density streets (coloured) ---
  add_line_layer(
    id         = "streets-density",
    source     = streets_dense,
    line_color = interpolate(
      column = "density",
      values = c(0, median(streets_dense$density), max(streets_dense$density)),
      stops  = c("#edf8e9", "#74c476", "#00441b")
    ),
    line_width   = interpolate(
      column = "density",
      values = c(0, max(streets_dense$density)),
      stops  = c(1.5, 6)
    ),
    line_opacity = 0.9,
    tooltip      = "street_label",
    popup        = concat(
      "<b>", get_column("street_label"), "</b><br/>",
      "Trees / 100m: ", number_format(get_column("density"), 1), "<br/>",
      "Total trees: ", get_column("total_trees")
    )
  ) %>%
  # --- 5. Numbered circles + labels (top) ---
  add_circle_layer(
    id                  = "top30-circles",
    source              = top30_sf,
    circle_radius       = 11,
    circle_color        = "white",
    circle_stroke_color = "#00441b",
    circle_stroke_width = 2,
    circle_opacity      = 0.95,
    tooltip             = "street_label",
    popup               = concat(
      "<b>#", get_column("rank"), " — ", get_column("street_label"), "</b><br/>",
      "Trees / 100m: ", number_format(get_column("density"), 1), "<br/>",
      "Total trees: ", get_column("total_trees"), "<br/>",
      "Length: ", number_format(get_column("total_length"), 0), "m"
    )
  ) %>%
  add_symbol_layer(
    id         = "top30-labels",
    source     = top30_sf,
    text_field = get_column("rank_label"),
    text_size  = 11,
    text_color = "#00441b",
    text_font  = list("Open Sans Bold")
  )
```

```{r}
# --- Save for downstream TSP analysis ---
st_write(top30_sf, here("data/processed/top30_streets.geojson"), delete_dsn = TRUE)
```
