---
title: "Vancouver Cherry Blossoms — Google Maps Route"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(sf)
library(dplyr)
library(spopt)
library(mapgl)
library(here)
```

```{r}
#| label: load-data
#| message: false

top30_stops <- st_read(here("data/processed/top30_streets.geojson"), quiet = TRUE)
tsp_stops   <- top30_stops %>% filter(rank <= 10)

# Helper: build a Google Maps directions URL from an ordered sf
google_maps_url <- function(stops_sf, address_col = "google_address") {
  stops_sf[[address_col]] %>%
    URLencode(reserved = TRUE) %>%
    paste(collapse = "/") %>%
    paste0("https://www.google.com/maps/dir/", .)
}
```

```{r}
#| label: tsp-euclidean

# Euclidean cost matrix as a suggested starting order for Google
coords_matrix        <- st_coordinates(tsp_stops)
cost_matrix_euclidean <- as.matrix(dist(coords_matrix))

result <- route_tsp(
  tsp_stops,
  start       = 1,
  end         = NULL,
  cost_matrix = cost_matrix_euclidean
)

stops_ordered <- result %>%
  filter(!is.na(.visit_order)) %>%
  arrange(.visit_order) %>%
  mutate(
    rank_label = as.character(rank),
    label      = as.character(.visit_order),
    lng        = st_coordinates(geometry)[, 1],
    lat        = st_coordinates(geometry)[, 2],
    # Append Vancouver, BC so Google resolves the address correctly
    google_address = paste0(street_label, ", Vancouver, BC")
  )
```

## Google Maps link

The stops below are pre-ordered using a straight-line TSP to give Google a reasonable starting sequence. Google will further optimize driving order — click **"Optimize order"** after opening the link.

```{r}
#| label: google-url
#| results: asis

google_url <- google_maps_url(stops_ordered)

cat(sprintf('<a href="%s" target="_blank" style="font-size:1.1em; font-weight:bold;">Open route in Google Maps ↗</a>\n\n', google_url))
```

## Stop list

```{r}
#| label: stop-table

stops_ordered %>%
  st_drop_geometry() %>%
  select(
    Stop            = .visit_order,
    Street          = street_label,
    `Trees / 100m`  = density,
    `Total trees`   = total_trees,
    `Lat, Lng`      = google_address
  ) %>%
  mutate(`Trees / 100m` = round(`Trees / 100m`, 1)) %>%
  knitr::kable()
```

## Map preview

```{r}
#| label: map-preview

all_streets <- st_read(here("data/final/streets.geojson"), quiet = TRUE)

top30_clusters <- top30_stops %>%
  st_transform(32610) %>%
  st_buffer(500) %>%
  st_union() %>%
  st_cast("POLYGON") %>%
  st_as_sf() %>%
  mutate(cluster_id = row_number()) %>%
  st_transform(4326)

# Straight lines connecting stops in TSP order for preview
route_preview <- stops_ordered %>%
  st_coordinates() %>%
  as.data.frame() %>%
  st_as_sf(coords = c("X", "Y"), crs = 4326) %>%
  st_combine() %>%
  st_cast("LINESTRING") %>%
  st_as_sf() %>%
  mutate(route = "preview")

walk_km    <- as.numeric(st_length(st_transform(route_preview, 32610))) / 1000
walk_miles <- walk_km * 0.621371
cat(sprintf("Walk route (top 10): %.1f km / %.1f miles (straight-line)\n", walk_km, walk_miles))

maplibre(style = openfreemap_style("bright"), bounds = top30_stops) %>%
  add_line_layer(
    id           = "all-streets",
    source       = all_streets,
    line_color   = "#dddddd",
    line_width   = 1,
    line_opacity = 0.3
  ) %>%
  add_fill_layer(
    id           = "clusters-fill",
    source       = top30_clusters,
    fill_color   = "#fde68a",
    fill_opacity = 0.25
  ) %>%
  add_line_layer(
    id           = "clusters-border",
    source       = top30_clusters,
    line_color   = "#d97706",
    line_width   = 1.5,
    line_opacity = 0.7
  ) %>%
  add_line_layer(
    id             = "route-preview",
    source         = route_preview,
    line_color     = "#16a34a",
    line_width     = 2.5,
    line_opacity   = 0.8,
    line_dasharray = c(2, 1)
  ) %>%
  add_circle_layer(
    id                  = "top30-circles",
    source              = top30_stops,
    circle_radius       = 10,
    circle_color        = "white",
    circle_stroke_color = "#d97706",
    circle_stroke_width = 1.5,
    circle_opacity      = 0.9,
    tooltip             = "street_label",
    popup               = concat(
      "<b>#", get_column("rank"), " — ", get_column("street_label"), "</b><br/>",
      "Trees / 100m: ", number_format(get_column("density"), 1), "<br/>",
      "Total trees: ", get_column("total_trees")
    )
  ) %>%
  add_symbol_layer(
    id         = "top30-labels",
    source     = top30_stops,
    text_field = get_column("rank_label"),
    text_size  = 10,
    text_color = "#d97706",
    text_font  = list("Open Sans Bold")
  ) %>%
  add_circle_layer(
    id                  = "tsp-stops",
    source              = stops_ordered,
    circle_radius       = 13,
    circle_color        = "white",
    circle_stroke_color = "#16a34a",
    circle_stroke_width = 2,
    circle_opacity      = 0.95,
    tooltip             = "street_label",
    popup               = concat(
      "<b>Stop ", get_column(".visit_order"), " — ", get_column("street_label"), "</b><br/>",
      "Trees / 100m: ", number_format(get_column("density"), 1), "<br/>",
      "Total trees: ", get_column("total_trees")
    )
  ) %>%
  add_symbol_layer(
    id         = "tsp-labels",
    source     = stops_ordered,
    text_field = get_column("label"),
    text_size  = 11,
    text_color = "#16a34a",
    text_font  = list("Open Sans Bold")
  )
```

---

## All 30 stops — bike route

For the ambitious cyclist. Google Maps limits route optimization to 10 stops at a time, so the 30 stops are split into 3 legs of 10. Open each leg in order and hit **Optimize route** on each one.

```{r}
#| label: tsp-bike
#| message: false

# TSP over all 30 to get a sensible geographic sequence
coords_all         <- st_coordinates(top30_stops)
cost_matrix_all    <- as.matrix(dist(coords_all))

result_all <- route_tsp(
  top30_stops,
  start       = 1,
  end         = NULL,
  cost_matrix = cost_matrix_all
)

bike_stops <- result_all %>%
  filter(!is.na(.visit_order)) %>%
  arrange(.visit_order) %>%
  mutate(
    label          = as.character(.visit_order),
    rank_label     = as.character(rank),
    google_address = paste0(street_label, ", Vancouver, BC"),
    leg            = ceiling(row_number() / 10)
  )
```

```{r}
#| label: bike-urls
#| results: asis

for (i in sort(unique(bike_stops$leg))) {
  leg_stops <- bike_stops %>% filter(leg == i)
  url       <- google_maps_url(leg_stops)
  cat(sprintf(
    '**Leg %d** (stops %d–%d): <a href="%s" target="_blank">Open in Google Maps ↗</a>\n\n',
    i,
    min(leg_stops$.visit_order),
    max(leg_stops$.visit_order),
    url
  ))
}
```

```{r}
#| label: bike-table

bike_stops %>%
  st_drop_geometry() %>%
  select(
    Stop           = .visit_order,
    Leg            = leg,
    Street         = street_label,
    `Trees / 100m` = density,
    `Total trees`  = total_trees
  ) %>%
  mutate(`Trees / 100m` = round(`Trees / 100m`, 1)) %>%
  knitr::kable()
```

```{r}
#| label: map-bike

bike_route_line <- bike_stops %>%
  st_coordinates() %>%
  as.data.frame() %>%
  st_as_sf(coords = c("X", "Y"), crs = 4326) %>%
  st_combine() %>%
  st_cast("LINESTRING") %>%
  st_as_sf() %>%
  mutate(route = "bike")

bike_km    <- as.numeric(st_length(st_transform(bike_route_line, 32610))) / 1000
bike_miles <- bike_km * 0.621371
cat(sprintf("Bike route (all 30): %.1f km / %.1f miles (straight-line)\n", bike_km, bike_miles))

maplibre(style = openfreemap_style("bright"), bounds = top30_stops) %>%
  add_line_layer(
    id           = "all-streets",
    source       = all_streets,
    line_color   = "#dddddd",
    line_width   = 1,
    line_opacity = 0.3
  ) %>%
  add_fill_layer(
    id           = "clusters-fill",
    source       = top30_clusters,
    fill_color   = "#fde68a",
    fill_opacity = 0.25
  ) %>%
  add_line_layer(
    id           = "clusters-border",
    source       = top30_clusters,
    line_color   = "#d97706",
    line_width   = 1.5,
    line_opacity = 0.7
  ) %>%
  add_line_layer(
    id             = "bike-route",
    source         = bike_route_line,
    line_color     = "#0ea5e9",
    line_width     = 2.5,
    line_opacity   = 0.8,
    line_dasharray = c(2, 1)
  ) %>%
  add_circle_layer(
    id                  = "bike-circles",
    source              = bike_stops,
    circle_radius       = 11,
    circle_color        = "white",
    circle_stroke_color = "#0ea5e9",
    circle_stroke_width = 2,
    circle_opacity      = 0.95,
    tooltip             = "street_label",
    popup               = concat(
      "<b>Stop ", get_column(".visit_order"), " (Leg ", get_column("leg"), ")</b><br/>",
      get_column("street_label"), "<br/>",
      "Trees / 100m: ", number_format(get_column("density"), 1)
    )
  ) %>%
  add_symbol_layer(
    id         = "bike-labels",
    source     = bike_stops,
    text_field = get_column("label"),
    text_size  = 10,
    text_color = "#0ea5e9",
    text_font  = list("Open Sans Bold")
  )
```
