---
title: "Vancouver Cherry Blossoms — Optimal Viewing Route"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(sf)
library(dplyr)
library(sfnetworks)
library(igraph)
library(tidygraph)
library(spopt)
library(mapgl)
library(here)
```

```{r}
#| label: load-data
#| message: false

top30_stops <- st_read(here("data/processed/top30_streets.geojson"), quiet = TRUE)
tsp_stops   <- top30_stops %>% filter(rank <= 10)
streets     <- st_read(here("data/final/streets.geojson"), quiet = TRUE)
```

## Build road network

```{r}
#| label: build-network

# Explode to individual linestrings so sfnetworks can build a clean graph
streets_lines <- streets %>%
  st_cast("MULTILINESTRING") %>%
  st_cast("LINESTRING")

# Build network, subdivide at intersections, smooth degree-2 nodes
net <- as_sfnetwork(streets_lines, directed = FALSE) %>%
  tidygraph::convert(to_spatial_subdivision) %>%
  tidygraph::convert(to_spatial_smooth) %>%
  activate("edges") %>%
  mutate(weight = edge_length())

# Keep only the largest connected component so no stop is unreachable
components   <- igraph::components(net)
largest_comp <- which.max(components$csize)

net_connected <- net %>%
  activate("nodes") %>%
  filter(components$membership == largest_comp)

# Snap each stop to its nearest node in the connected network
network_nodes <- net_connected %>% activate("nodes") %>% st_as_sf()
nearest_nodes <- st_nearest_feature(tsp_stops, network_nodes)

cat(sprintf("Network: %d nodes, %d edges\n",
  igraph::vcount(net_connected),
  igraph::ecount(net_connected)
))
```

## Build cost matrix

Network distances (metres) between all pairs of stops via road.

```{r}
#| label: cost-matrix

cost_matrix <- igraph::distances(
  net_connected,
  v       = nearest_nodes,
  to      = nearest_nodes,
  weights = igraph::E(net_connected)$weight
)

n_inf <- sum(is.infinite(cost_matrix))
if (n_inf > 0) {
  warning(sprintf("%d unreachable stop pair(s) — check network connectivity.", n_inf))
} else {
  message("Cost matrix OK — all stops reachable.")
}
```

## Solve TSP

Open route: starts at stop 1, does not need to return to start.

```{r}
#| label: solve-tsp

result <- route_tsp(
  tsp_stops,
  start       = 1,
  end         = NULL,
  cost_matrix = cost_matrix
)

meta <- attr(result, "spopt")
cat(sprintf("Optimized route: %.0f m (%.1f km)\n", meta$total_cost, meta$total_cost / 1000))
cat(sprintf("Nearest-neighbor baseline: %.0f m\n", meta$nn_cost))
cat(sprintf("Improvement: %.1f%%\n", meta$improvement_pct))
```

## Visit order

```{r}
#| label: visit-order

result_ordered <- result %>%
  filter(!is.na(.visit_order)) %>%
  arrange(.visit_order) %>%
  mutate(
    label    = as.character(.visit_order),
    stop_idx = row_number()
  )

result_ordered %>%
  st_drop_geometry() %>%
  select(
    Stop          = .visit_order,
    Street        = street_label,
    `Trees/100m`  = density,
    `Total trees` = total_trees
  ) %>%
  mutate(`Trees/100m` = round(`Trees/100m`, 1)) %>%
  knitr::kable()
```

## Route map

```{r}
#| label: route-geometry

# Extract edges sf once outside the loop to avoid repeated activation
edges_sf <- net_connected %>% activate("edges") %>% st_as_sf()

ordered_nodes <- nearest_nodes[result_ordered$stop_idx]

route_legs <- lapply(seq_len(nrow(result_ordered) - 1), function(i) {
  path <- igraph::shortest_paths(
    net_connected,
    from    = ordered_nodes[i],
    to      = ordered_nodes[i + 1],
    weights = igraph::E(net_connected)$weight
  )$vpath[[1]]

  if (length(path) < 2) return(NULL)

  edge_ids <- igraph::get_edge_ids(
    net_connected,
    vp = as.vector(rbind(
      head(as.integer(path), -1),
      tail(as.integer(path), -1)
    ))
  )

  # get_edge_ids returns 0 for unmatched pairs — drop them
  edge_ids <- edge_ids[edge_ids > 0]
  if (length(edge_ids) == 0) return(NULL)

  edges_sf %>%
    slice(edge_ids) %>%
    st_union() %>%
    st_as_sf() %>%
    mutate(leg = i) %>%
    rename(geometry = x)
})

route_line <- bind_rows(Filter(Negate(is.null), route_legs))
```

```{r}
#| label: map-route

all_streets <- st_read(here("data/final/streets.geojson"), quiet = TRUE)

# Cluster blobs from all 30 stops
top30_clusters <- top30_stops %>%
  st_transform(32610) %>%
  st_buffer(500) %>%
  st_union() %>%
  st_cast("POLYGON") %>%
  st_as_sf() %>%
  mutate(cluster_id = row_number()) %>%
  st_transform(4326)

maplibre(style = openfreemap_style("bright"), bounds = top30_stops) %>%
  # Background streets
  add_line_layer(
    id           = "all-streets",
    source       = all_streets,
    line_color   = "#dddddd",
    line_width   = 1,
    line_opacity = 0.3
  ) %>%
  # Cluster blobs (top 30)
  add_fill_layer(
    id           = "clusters-fill",
    source       = top30_clusters,
    fill_color   = "#fde68a",
    fill_opacity = 0.25
  ) %>%
  add_line_layer(
    id           = "clusters-border",
    source       = top30_clusters,
    line_color   = "#d97706",
    line_width   = 1.5,
    line_opacity = 0.7
  ) %>%
  # Road-following route (top 10 only)
  add_line_layer(
    id             = "tsp-route",
    source         = route_line,
    line_color     = "#2563eb",
    line_width     = 3,
    line_opacity   = 0.8,
    line_dasharray = c(2, 1)
  ) %>%
  # Top 30 circles (all locations, smaller)
  add_circle_layer(
    id                  = "top30-circles",
    source              = top30_stops,
    circle_radius       = 10,
    circle_color        = "white",
    circle_stroke_color = "#d97706",
    circle_stroke_width = 1.5,
    circle_opacity      = 0.9,
    tooltip             = "street_label",
    popup               = concat(
      "<b>#", get_column("rank"), " — ", get_column("street_label"), "</b><br/>",
      "Trees / 100m: ", number_format(get_column("density"), 1), "<br/>",
      "Total trees: ", get_column("total_trees")
    )
  ) %>%
  add_symbol_layer(
    id         = "top30-labels",
    source     = top30_stops,
    text_field = get_column("rank_label"),
    text_size  = 10,
    text_color = "#d97706",
    text_font  = list("Open Sans Bold")
  ) %>%
  # Top 10 route stops on top (larger, blue)
  add_circle_layer(
    id                  = "tsp-stops",
    source              = result_ordered,
    circle_radius       = 13,
    circle_color        = "white",
    circle_stroke_color = "#2563eb",
    circle_stroke_width = 2,
    circle_opacity      = 0.95,
    tooltip             = "street_label",
    popup               = concat(
      "<b>Stop ", get_column(".visit_order"), " — ", get_column("street_label"), "</b><br/>",
      "Trees / 100m: ", number_format(get_column("density"), 1), "<br/>",
      "Total trees: ", get_column("total_trees")
    )
  ) %>%
  add_symbol_layer(
    id         = "tsp-labels",
    source     = result_ordered,
    text_field = get_column("label"),
    text_size  = 11,
    text_color = "#2563eb",
    text_font  = list("Open Sans Bold")
  )
```

## Straight-line TSP (simpler alternative)

Same top 10 stops, but the cost matrix uses straight-line (Euclidean) distances instead of road network distances. Faster to compute and avoids road network issues — useful as a sanity check or when the road-following route looks wrong.

```{r}
#| label: solve-tsp-euclidean

coords_matrix <- st_coordinates(tsp_stops)
cost_matrix_euclidean <- as.matrix(dist(coords_matrix))

result_eu <- route_tsp(
  tsp_stops,
  start       = 1,
  end         = NULL,
  cost_matrix = cost_matrix_euclidean
)

meta_eu <- attr(result_eu, "spopt")
cat(sprintf("Optimized route: %.0f m (%.1f km)\n", meta_eu$total_cost, meta_eu$total_cost / 1000))
cat(sprintf("Nearest-neighbor baseline: %.0f m\n", meta_eu$nn_cost))
cat(sprintf("Improvement: %.1f%%\n", meta_eu$improvement_pct))
```

```{r}
#| label: visit-order-euclidean

result_eu_ordered <- result_eu %>%
  filter(!is.na(.visit_order)) %>%
  arrange(.visit_order) %>%
  mutate(
    label    = as.character(.visit_order),
    stop_idx = row_number()
  )

result_eu_ordered %>%
  st_drop_geometry() %>%
  select(
    Stop          = .visit_order,
    Street        = street_label,
    `Trees/100m`  = density,
    `Total trees` = total_trees
  ) %>%
  mutate(`Trees/100m` = round(`Trees/100m`, 1)) %>%
  knitr::kable()
```

```{r}
#| label: map-route-euclidean

# Straight lines connecting stops in visit order
route_straight <- result_eu_ordered %>%
  st_coordinates() %>%
  as.data.frame() %>%
  st_as_sf(coords = c("X", "Y"), crs = 4326) %>%
  st_combine() %>%
  st_cast("LINESTRING") %>%
  st_as_sf() %>%
  mutate(route = "straight")

maplibre(style = openfreemap_style("bright"), bounds = top30_stops) %>%
  # Background streets
  add_line_layer(
    id           = "all-streets",
    source       = all_streets,
    line_color   = "#dddddd",
    line_width   = 1,
    line_opacity = 0.3
  ) %>%
  # Cluster blobs (top 30)
  add_fill_layer(
    id           = "clusters-fill",
    source       = top30_clusters,
    fill_color   = "#fde68a",
    fill_opacity = 0.25
  ) %>%
  add_line_layer(
    id           = "clusters-border",
    source       = top30_clusters,
    line_color   = "#d97706",
    line_width   = 1.5,
    line_opacity = 0.7
  ) %>%
  # Straight-line route
  add_line_layer(
    id             = "tsp-route-straight",
    source         = route_straight,
    line_color     = "#9333ea",
    line_width     = 2.5,
    line_opacity   = 0.8,
    line_dasharray = c(2, 1)
  ) %>%
  # Top 30 circles
  add_circle_layer(
    id                  = "top30-circles",
    source              = top30_stops,
    circle_radius       = 10,
    circle_color        = "white",
    circle_stroke_color = "#d97706",
    circle_stroke_width = 1.5,
    circle_opacity      = 0.9,
    tooltip             = "street_label",
    popup               = concat(
      "<b>#", get_column("rank"), " — ", get_column("street_label"), "</b><br/>",
      "Trees / 100m: ", number_format(get_column("density"), 1), "<br/>",
      "Total trees: ", get_column("total_trees")
    )
  ) %>%
  add_symbol_layer(
    id         = "top30-labels",
    source     = top30_stops,
    text_field = get_column("rank_label"),
    text_size  = 10,
    text_color = "#d97706",
    text_font  = list("Open Sans Bold")
  ) %>%
  # Top 10 route stops
  add_circle_layer(
    id                  = "tsp-stops",
    source              = result_eu_ordered,
    circle_radius       = 13,
    circle_color        = "white",
    circle_stroke_color = "#9333ea",
    circle_stroke_width = 2,
    circle_opacity      = 0.95,
    tooltip             = "street_label",
    popup               = concat(
      "<b>Stop ", get_column(".visit_order"), " — ", get_column("street_label"), "</b><br/>",
      "Trees / 100m: ", number_format(get_column("density"), 1), "<br/>",
      "Total trees: ", get_column("total_trees")
    )
  ) %>%
  add_symbol_layer(
    id         = "tsp-labels",
    source     = result_eu_ordered,
    text_field = get_column("label"),
    text_size  = 11,
    text_color = "#9333ea",
    text_font  = list("Open Sans Bold")
  )
```
